XY: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-03-27 paper/bare_conf.tex
added some paragraphs

@@

The complexity of Firedrake-generated kernels depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. The stiffness matrix in the code is called $A$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest. This is not true in general, since bent elements might be used in the unstructured mesh, which would require the Jacobian be re-computed at every $i$ iteration; 2) memory accesses along the three loop dimensions are always stride-1; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) arrays like $X$, $Y$, ..., which represent so called basis functions and their derivatives, are constants; 6) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
%All the work done in the polyhedral model should be useful, but our kernels fit L1, so they need special attention for what concerns key aspects like managing registers, exploiting ILP in its various forms, and reducing loop overhead. 
%Follows a description of code transformations using our compiler-driven approach.

>>>>

\begin{algorithm}[t]
\small
\KwIn{stiffness matrix A (initialized to 0),\\~~~~~~~~~element coordinates in the domain (array),\\~~~~~~~~~(optional) fields (e.g. velocity, array)}
\KwOut{stiffness matrix A (computed)}
\nlset{1}~~- Jacobian computation, using element coordinates \\
~~- Declaration of constant data, among which basis function\\
~~- Computation of the stiffness matrix
\caption{General structure of a local assembly kernel generated by Firedrake.}
\label{code:general-structure}
\end{algorithm}

@@

The general structure of a Firedrake-generated kernel is shown in Figure~\ref{code:general-structure}. Input is a pointer to a 0-valued bi-dimensional array, which will be used to store the stiffness matrix; a pointer to the element coordinates in the discretized domain; one or more pointers to so called fields, for instance indicating value of velocity or pressure in the element. The output is the evaluated element stiffness matrix. A local assembly kernel can be logically split into three parts: 1) calculation of the Jacobian matrix, its determinant and its inverse (referred to as $det$ and $K$ in the following code snippets), 2) declaration of constant arrays (identified by letters $X$, $Y$, ...), 3) evaluation of the stiffness matrix (called $A$) in an affine loop nest. In finite element terminology, constant (bi-dimensional) arrays are called ``basis functions'' (or derivatives of basis functions). They are identical to all mesh elements.

The actual complexity of a local assembly kernel depends on the finite element problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of the stiffness matrix can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that the space of codes that Firedrake can generate is infinite, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest. This is not true in general, since bent elements might be used in the unstructured mesh, which would require the Jacobian be re-computed at every $i$ iteration; 2) memory accesses along the three loop dimensions are always stride-1; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the $ijk$ loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumnstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that its execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.
